---
title: 'DSA 8070 Project I'
author: "Alexander Harriman"
date: 'Due: October 19, 2023'
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing Financial Data

### Load the Financial Data using the quantmod R package

Let's consider the S&P 500, Walmart, Apple, and IBM stocks in this project.

```{r,message=FALSE, warning = FALSE}
library("quantmod")
getSymbols("^GSPC", src = "yahoo") # S&P500 data.
getSymbols("WMT", src = "yahoo") # Walmart data.
getSymbols("AAPL", src = "yahoo") # Apple data.
getSymbols("IBM", src = "yahoo") # IBM data.
```

### Calculate the week return

The code chunk below walks you through the calculation of the log return. Let $Y_{t}$ represent the stock price at time $t$, then the log return, denoted by $r_{t}$, is defined as:

$$\log(\frac{Y_{t}}{Y_{t-1}}) = \log(Y_{t})-\log(Y_{t-1}).$$


```{r}
# extract time information
id <- seq(1, 4206, 5)
date <- as.Date(time(AAPL)[id])
year <- format(date, "%Y")
month <- format(date, "%m")
# grab daily closing prices on a particular weekday 
apple <- AAPL$AAPL.Close[id]
ibm <- IBM$IBM.Close[id]
wmt <- WMT$WMT.Close[id]
sp500 <- GSPC$GSPC.Close[id]
closing <- data.frame(Apple = apple, IBM = ibm, 
                      SP500 = sp500, WMT = wmt)
# calculate the week returns
return <- apply(closing, 2, function(x) diff(log(x)))
```



### Question 1

Please conduct an exploratory analysis and summarize your findings. Keep in mind that these are time series data, so you should also investigate their time-dependent characteristics.

**Code**

```{r, warning = FALSE}
return <- as.data.frame(return)
return$Time <- as.POSIXct(rownames(return))

# Plot a GGPairs plot
library(GGally)
library(ggplot2)
stocks <- ggpairs(return[,-5]) + theme_bw()
stocks


#Find Means and Variances for Each Stock
stockMean <- apply(return[,1:4], 2, mean)
stockVar <- apply(return[,1:4], 2, var)


stockMean
stockVar


#Find the Correlation Matrix
library(ggcorrplot)

corr <- round(cor(return[,1:4]), 1)
# Visualize
ggcorrplot(corr, p.mat = cor_pmat(return[,1:4]),
           hc.order = TRUE, type = "lower",
           color = c("#FC4E07", "white", "#00AFBB"),
           outline.col = "white", lab = TRUE)
```

Time-Dependent Analysis:

```{r}
#Apple
ggplot(return, aes(x = Time, y = AAPL.Close)) +
  geom_point() +
  ggtitle('Week Return Values of Apple Stock over Time') +
  xlab('Time') +
  ylab('Week Return, $') +
  geom_smooth(method = 'lm')

#IBM
ggplot(return, aes(x = Time, y = IBM.Close)) +
  geom_point() +
  ggtitle('Week Return Values of IBM Stock over Time') +
  xlab('Time') +
  ylab('Week Return, $') +
  geom_smooth(method = 'lm')

#S&P 500
ggplot(return, aes(x = Time, y = GSPC.Close)) +
  geom_point() +
  ggtitle('Week Return Values of the S&P 500 Exchange over Time') +
  xlab('Time') +
  ylab('Week Return, $') +
  geom_smooth(method = 'lm')

#Walmart
ggplot(return, aes(x = Time, y = WMT.Close)) +
  geom_point() +
  ggtitle('Week Return Values of Walmart Stock over Time') +
  xlab('Time') +
  ylab('Week Return, $') +
  geom_smooth(method = 'lm')
```

```{r, warning = FALSE}
#Checking Stock Closing Values Using Auto-ARIMA
library(forecast)
appleArima <- auto.arima(closing$AAPL.Close)

summary(appleArima)
futureApple <- forecast(appleArima, 50)
plot(futureApple)


ibmArima <- auto.arima(closing$IBM.Close)

summary(ibmArima)
futureIBM <- forecast(ibmArima, 50)
plot(futureIBM)


walArima <- auto.arima(closing$WMT.Close)

summary(walArima)
futureWal <- forecast(walArima, 50)
plot(futureWal)


spArima <- auto.arima(closing$GSPC.Close)

summary(spArima)
futureSP <- forecast(spArima, 50)
plot(futureSP)
```

```{r, warning = FALSE}
#Checking Weekly Returns Using Auto-ARIMA
library(forecast)
appleArima <- auto.arima(return$AAPL.Close)

summary(appleArima)
futureApple <- forecast(appleArima, 50)
plot(futureApple)


ibmArima <- auto.arima(return$IBM.Close)

summary(ibmArima)
futureIBM <- forecast(ibmArima, 50)
plot(futureIBM)


walArima <- auto.arima(return$WMT.Close)

summary(walArima)
futureWal <- forecast(walArima, 50)
plot(futureWal)


spArima <- auto.arima(return$GSPC.Close)

summary(spArima)
futureSP <- forecast(spArima, 50)
plot(futureSP)
```



**Conclusion**

**All of the stocks have a mean week return generally around $0, with all of them being slightly positive. The variances are also all very similar and close to 0.**

**Apple and IBM stocks are both highly correlated with the S&P 500 stock value, while Walmart's correlation is much less clear. The correlations between individual company stocks are fairly low, but still significantly positive. Apple and IBM have the highest correlations, which makes sense as both are in the tech industry and would see rises and falls together.**

**Over time, none of the stocks have adjusted too far from a week return of $0. Apple and IBM stocks appear to have slightly declined over the years, while the S&P 500 slightly increased and Walmart remained relatively steady.**



**All of the raw stock values conform to an ARIMA model. Apple and IBM do not have any AR or MA effects, both having ARIMA(0,1,0) modeling. Walmart adds an ARMA(2,2) to the I = 1 value, while S&P 500 uses an ARIMA(0,1,1) model.**

**Outside of Apple being an ARMA(2,2) model, none of the weekly returns have a significant autocorrelation effect. IBM weekly returns are white noise, while Walmart and IBM are MA(1) models.**

### Question 2

Perform a hypothesis test at $\alpha= 0.05$ to determine whether there is evidence to support the claim that the weekly returns for S&P 500, Walmart, Apple, and IBM stocks are not all equal to 0.

**Code**

```{r}
#Check for need to normalize data
head(return)
tail(return)

#The values are all relatively close to each other, so normalization is not needed.

#Find the Means for Each Stock

stockMeans <- colMeans(return[,1:4]) #Since column 5 is Time
stockMeans

#Find the Covariance Matrix
stockCov <- cov(return[,1:4])
stockCov

#Find Number of Datapoints and Number of Variables

n <- dim(return)[1]
#Ignore the Time variable
p <- dim(return[,1:4])[2]


#mu0: all means are equal to 0
mu0 <- rep(0,4)


#Find the Hotelling T-Squared value
stockT <- as.numeric(n * t(stockMeans - mu0) %*% solve(stockCov) %*% (stockMeans - mu0))


#Find the F value
(stockF <- stockT * ((n - p) / ((n - 1) * p)))

pf(stockF, p, n-p, lower.tail = FALSE)
```

**Conclusion**

**The F statistic is 3.132, which results in a p-value of 0.014. Based on an alpha of 0.05, the null hypothesis is rejected, and we conclude at the 95% confidence level that there is a significant difference in returns between at least one pair of stocks.**


### Question 3

Select two non-overlapping, consecutive years (e.g., 2007-2008 and 2017-2018) to conduct a hypothesis test at $\alpha= 0.05$ to determine whether there is evidence supporting the claim that the weekly returns for S&P 500, Walmart, Apple, and IBM stocks are not all equal during these two periods.

**Code**

```{r}
#First Pair: 2009-2010
firstPair <- return[return$Time >= '2009-01-01' & return$Time <= '2010-12-31',]

#Second Pair: 2016-2017
secondPair <- return[return$Time >= '2016-01-01' & return$Time <= '2017-12-31',]

totalPair <- rbind(firstPair, secondPair)


stockMeans <- colMeans(totalPair[,1:4]) #Since column 5 is Time
stockMeans

#Find the Covariance Matrix
stockCov <- cov(return[,1:4])
stockCov

#Find Number of Datapoints and Number of Variables

n <- dim(return)[1]
#Ignore the Time variable
p <- dim(return[,1:4])[2]


#mu0: all means are equal to 0
mu0 <- rep(0,4)


#Find the Hotelling T-Squared value
stockT <- as.numeric(n * t(stockMeans - mu0) %*% solve(stockCov) %*% (stockMeans - mu0))


#Find the F value
(stockF <- stockT * ((n - p) / ((n - 1) * p)))

pf(stockF, p, n-p, lower.tail = FALSE)
```

**Conclusion**


**The F-statistic is 9.539, which results in a p-value of 1.52E-7. Since this p-value is less than 0.05, the null hypothesis is rejected. Thus, we conclude at a 95% confidence level that at least one of the stocks are different when focusing on these two periods of data.**

### Question 4

Draw a 95% confidence ellipsoid for the mean vector of weekly returns for Apple and IBM (or any other pair of stocks).

**Code**

```{r, warning = FALSE}
#Select specific stocks: IBM and Walmart

twoStock <- return[2] |>
  cbind(return[4])

head(twoStock)

p <- 2

stockCor <- sqrt(((n - 1) * p / (n - p)) * qf(0.95, p, n) / qchisq(0.95, p)) 




#Find the correlation coefficient for the two stocks
rho = cor(twoStock)[1,2]
#Find the sample variance for the two stocks
twoVar <- apply(twoStock, 2, var)
#Find the sample mean for the two stocks
twoMean <- apply(twoStock, 2, mean)



##Using Simultaneous 95% CIs
#Find the multiplier

alpha <- 0.05

multiplier <- sqrt((p * (n - 1) / (n - p)) * qf(1 - alpha, p, n - p))

#The qf function should be used to get the F statistic value in a simulatenous CI multiplier

(CIMin <- twoMean[1] + c(-1, 1) * multiplier * (twoVar[1] / sqrt(n)))
## mu2
(CIMax <- twoMean[2] + c(-1, 1) * multiplier * (twoVar[2] / sqrt(n)))

library(ellipse)
par(las = 1, mgp = c(2, 1, 0), mar = c(3.5, 3.5, 0.6, 0.6))
plot(ellipse(rho, scale = stockCor * twoVar / sqrt(n), centre = twoMean), type = 'l',
las = 1, bty = "n", xaxt = "n", yaxt = "n",
xlim = range(CIMin),
ylim = range(CIMax) * c(0.995, 1.025), xlab = expression(bar(x)[1]),
ylab = expression(bar(x)[2]))
points(twoMean[1], twoMean[2], pch = "+")
xg <- seq(twoMean[1] - 3 * (twoVar[1] / sqrt(n)), twoMean[1] + 3 * (twoVar[1] / sqrt(n)), twoVar[1] / sqrt(n))
yg <- seq(twoMean[2] - 3 * (twoVar[2] / sqrt(n)), twoMean[2] + 3 * (twoVar[2] / sqrt(n)), twoVar[2] / sqrt(n))
axis(1, at = xg)
axis(2, at = yg)

rect(CIMin[1], CIMin[2], CIMax[1], CIMax[2], border = "green", lwd = 0.5)

```

### Question 5

Describe how a copula modeling approach can be used to model the joint distribution of weekly returns for Apple and IBM (or any other pair of stocks).

**Answer: By taking the percentile values of the stocks and modeling them into a [0,1] space, the joint distribution can become similar to a multivariate normal distribution regardless of whether or not the stocks themselves are normally distributed. This also allows the model to have non-normal marginal distributions, allowing for more options for modeling the weekly returns' dependence on each other.**

Try implementing copula modeling for the joint distribution of weekly returns for Apple and IBM (or any other pair of stocks) in R.

**Code**

```{r}
library(VineCopula)

#IBM and Walmart Stocks will be carried over from above problem

plot(twoStock, pch = 16, cex = 0.8, 
     xlab = "IBM", ylab = "Walmart", main = 'IBM and Walmart Weekly Returns')

#Get the Percentiles to Get Best Copula
stockPerc <- apply(twoStock, 2, function(x) rank(x) / (length(x) + 1))


#Find the Best Copula
copStock <- BiCopSelect(stockPerc[,1], stockPerc[,2], familyset = NA)

copStock

#Since t was chosen, a Student t copula will be used.

#Get simulated values from the copula
tSim <- BiCopEst(stockPerc[,1], stockPerc[,2], family = 2)


#Plot the Copula
plot(tSim)

summary(tSim)
```


### Question 6


Perform a multivariate linear regression that uses SP500 as the predictor to model Walmart, Apple, and IBM stocks.

**Code**

```{r}
#Using the raw closing data instead of the weekly returns

#Get the three stocks into one data frame to use as Y

depStocks <- closing[1:2] |>
  cbind(closing[4])

depStocks <- as.matrix(depStocks)

stockLM <- lm(depStocks ~ GSPC.Close, data = closing)
summary(stockLM)
```

**Conclusion: S&P 500 stocks can best model Walmart stocks, as when Walmart is the dependent variable of choice, the Adjusted R^2 is highest. For each $1 increase in S&P 500 stock prices, Walmart increases by about 3 cents.**

**Apple stocks are also well-predicted by S&P 500, as a $1 increase in the S&P 500 sees a 5 cent rise in Apple stocks. However, there is little predictive power with the IBM stocks.**



```{r}
#Using the the weekly returns

#Get the three stocks into one data frame to use as Y

depStocks <- return[1:2] |>
  cbind(return[4])

depStocks <- as.matrix(depStocks)

stockLM <- lm(depStocks ~ GSPC.Close, data = return)
summary(stockLM)
```

**Conclusion: S&P 500 weekly returns are able to significantly predict the three stocks, but none of the three models have a very high adjusted R^2 value. For each $1 increase in the S&P 500 weekly return, the Apple weekly return increases by $1.11, the IBM weekly return increases by $0.84, and the Walmart weekly return (the weakest of the three stocks) increases by $0.48.**
